plyranges: a grammar for transforming genomics data

### 2018-03-22 paper.Rmd
reshuffling

By definition the _Ranges_ class encapsulate tidy data prinicples because
every row corresponds to a sequence and every column corresponds to a variable.
Hence, we have designed the `plyranges` DSL to have the grammar
and principles of `dplyr`: fluency, consistency, cohesion and endomorphism.

We have designed the API to be fluent. Every function call corresponds to an
action on a _Ranges_ object (they are named verbs) and where possible
functions have few arguments. Each verb is constructed
to enable a tab completion based workflow. Both of these aspects reduce the 
cognitive load on a new user since most manipulations can be performed with a 
vocabulary of several verbs, rather than having to memorise functions with many 
arguments that are nouns (as is required in the existing Bioconductor packages). 
This is also has the advantage of allowing users to write human readable code
because verbs describe what the code is doing rather than how its doing it.

Workflows can be composed by chaining verbs together via the forward pipe operator,`%>%` 
(exported from the R package _magrittr_ @R-magrittr). This is possible because every 
function call is an endomorphism: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
means that computations are performed and evaluated in the context of the 
_Ranges_ objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with _plyranges_ becomes more difficult 
because a user needs to capture expressions inside function calls and then
unquote them.
<!--- need to make this clearer? --->

>>>>

By definition the _Ranges_ class encapsulate tidy data prinicples because
every row corresponds to a sequence and every column corresponds to a variable.
Hence, we have designed the `plyranges` DSL to have the grammar
and design principles of `dplyr`: fluency, consistency, cohesion and endomorphism.

To promote fluency we have designed every function call to perform an
action on a _Ranges_ object and where possible
functions have few arguments. Each function is named after a verb and is designed
to enable a tab completion based workflow. Both of these aspects reduce the 
cognitive load on a new user since most manipulations can be performed with a 
vocabulary of several verbs, rather than having to memorise functions (that
are named as nouns) with many arguments (which is required in the existing Bioconductor packages). Workflows can be composed by chaining verbs together into
'sentences' via the forward pipe operator,`%>%` (exported from the R package `magrittr` @R-magrittr). Overall, this allows users to write human readable code
because workflows describe what the code is doing rather than how its doing it.

In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
means that computations are performed and evaluated in the context of the 
_Ranges_ objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with _plyranges_ becomes more difficult 
because a user needs to capture expressions inside function calls and then
unquote them.

 This is possible because every 
function call is an endomorphism: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

<!--- need to make this clearer? --->
